---
title: "Introduction to sofamatic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to sofamatic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

> *Automatic SOFA scoring for R.*

## What this package does

`sofamatic` computes the Sequential Organ Failure Assessment (SOFA) score
and its six sub-scores from clinical data. It exposes one vectorised
function per component plus a `calculate_sofa()` aggregator that takes a
long-format data frame and returns it augmented with all sub-scores, the
total score, and (optionally) longitudinal helpers (`SOFA_admission`,
`SOFA_max`).

The thresholds follow the canonical definition (Vincent et al., 1996).
Both SI units (kPa, µmol/L) and conventional units (mmHg, mg/dL) are
supported.

```{r setup}
library(sofamatic)
```

## Single-component scoring

Each component is a vectorised function:

```{r}
score_respiratory(c(450, 350, 250, 150, 80),
                  ventilated = c(0, 0, 1, 1, 1),
                  units = "mmHg")

score_coagulation(c(180, 130, 75, 35, 12))

score_liver(c(0.8, 1.5, 3, 8, 14), units = "mg/dL")

score_cardiovascular(map = c(80, 65, 60, 60),
                     norepinephrine = c(0, 0, 0.05, 0.2))

score_neurological(c(15, 13, 11, 8, 4))

score_renal(creatinine = c(0.9, 1.5, 2.5, 4, 6), units = "mg/dL")
```

Use `?score_respiratory` (etc.) to see the full thresholds and unit
conventions.

## Cross-sectional aggregation

`calculate_sofa()` accepts a data frame and column names. Defaults match
the parameter names (`pao2_fio2`, `platelets`, `bilirubin`, `map`, `gcs`,
`creatinine`, …); every column name is overridable.

```{r}
df <- example_sofa_data(n_subjects = 4, days_each = 1)
calculate_sofa(df)[, c("SOFA_resp", "SOFA_coag", "SOFA_liver",
                       "SOFA_cardio", "SOFA_neuro", "SOFA_renal",
                       "SOFA_score")]
```

## Longitudinal data

For long-format longitudinal data (one row per subject per day), supply
`id` and `time` to also get `SOFA_admission` (first day) and `SOFA_max`
(worst day) per subject.

```{r}
df  <- example_sofa_data(n_subjects = 3, days_each = 4)
out <- calculate_sofa(df, id = "subject_id", time = "day")
out[, c("subject_id", "day",
        "SOFA_score", "SOFA_admission", "SOFA_max")]
```

## Handling missing data

`calculate_sofa(na_strategy = "sum_na_zero")` (the default) treats `NA`
sub-scores as 0 in the total — convenient when only some components are
unmeasured. Use `na_strategy = "propagate"` to instead return `NA`
whenever any sub-score is missing.

For the neurological sub-score, `score_neurological()` accepts a
`sedation` flag. A missing GCS in a non-sedated patient is treated as
"alert" (score 0); a missing GCS in a sedated patient is treated as
moderate impairment (score 2). This reflects the common ICU practice in
which fully responsive, non-sedated patients are often not re-assessed,
leaving the GCS field blank.

## The synthetic example dataset

`example_sofa_data()` returns a fully synthetic, longitudinal ICU-style
dataset. The data are simulated and never describe a real patient.
Distributions are loosely calibrated to look plausible (e.g., lower
PaO2/FiO2 in ventilated rows, occasional missing GCS in sedated rows,
~10 % dialysis prevalence), but they should never be used to draw
clinical conclusions.

```{r}
str(example_sofa_data(n_subjects = 2, days_each = 2))
```

## Units and thresholds at a glance

| Component       | Threshold variable | Standard unit | Conventional unit |
|-----------------|--------------------|--------------|--------------------|
| Respiratory     | PaO2/FiO2          | kPa          | mmHg               |
| Coagulation     | Platelets          | 10⁹/L        | —                  |
| Liver           | Bilirubin          | µmol/L       | mg/dL              |
| Cardiovascular  | MAP, vasoactives   | mmHg, µg/kg/min | —               |
| Neurological    | GCS                | 3–15         | —                  |
| Renal           | Creatinine, urine  | µmol/L, mL/24h  | mg/dL           |

## References

Vincent, J. L., Moreno, R., Takala, J., Willatts, S., De Mendonça, A.,
Bruining, H., … Thijs, L. G. (1996). The SOFA (Sepsis-related Organ
Failure Assessment) score to describe organ dysfunction/failure.
*Intensive Care Medicine*, 22(7), 707–710.
<https://doi.org/10.1007/BF01709751>
